### 2022-09-15 reports/paper/paper.Rmd
Another draft of  manuscript

---
# fontsize: 11pt
# mainfont: arial
output:
  # pdf_document:
  #   fig_caption: true
  #   latex_engine: lualatex
  #   includes:
  #     in_header: "~/HeadRs/Stathead.sty"
  #   keep_tex: true
  bookdown::word_document2:
    number_sections: false
    global_numbering: true
    reference_docx: "~/HeadRs/rmd_templates/style-guide-arial-11.docx"
bibliography: "`r path.expand('~/Box Sync/papers/EnvEpi.bib')`"
csl: "`r path.expand('~/HeadRs/csl/ama.csl')`"
---

@@

**Manuscript title:** Evaluating hypothetical limits on metalworking fluid exposure for reducing non-Hodgkin lymphoma incidence: An application of the hazard-extended iterative conditional expectation parametric g-formula

**Authors:** Kevin T. Chen,^a,b^, Sally Picciotto^b^, Ellen A. Eisen^b^

@@

**Data and computing code:**

@@

# Abstract (250 words)

**Background:** The second half of the 20th century saw a fourfold rise in non-Hodgkin lymphoma (NHL) incidence. Over the same period, concerns about the carcinogenicity of oil-based metalworking fluids (MWF) in automobile manufacturing fueled a transition to water-based soluble MWF. Soluble MWF, a complex mixture of diverse chemical additives, has become the most widely used class of MWF in the industry. We estimated the effect of limiting annual exposure to soluble MWF on NHL incidence in the United Auto Workers-General Motors (UAW-GM) Cohort Study (n = 33,138). We based our hypothetical interventions on the NIOSH recommended exposure limit (REL) of 0.5 mg/m3 for total MWF particulate mass.

**Methods:** We estimated counterfactual cumulative incidence of NHL from 1985 to 2004 in the UAW-GM cohort of hourly autoworkers ever employed at 3 Michigan plants for at least 3 years. Using the hazard-extended iterative conditional expectation g-computation, we adjusted for time-varying confounding and censoring affected by prior exposure. We contrasted counterfactual cumulative incidence under no intervention on exposure to when annual soluble exposure was capped at the REL, half the REL, and a tenth of the REL. Confidence intervals were computed using the nonparametric bootstrap.

Results: There were 211 NHL cases during the follow-up period. Lower annual soluble exposure resulted in lower cumulative incidence of NHL. The risk ratio when capping soluble exposure at the REL compared to no intervention was 0.90 (95% CI: 0.78, 1.03). The corresponding difference in cumulative incidence was -28 cases (95%CI: -63, 8 cases). Capping soluble exposure at a tenth of the REL resulted in a risk ratio of 0.84 (95% CI: 0.68, 1.00) and a difference in cumulative incidence of -45 cases (95%CI: -88, 1 case).
Conclusion: The current REL was based on evidence linking MWF exposure to respiratory and skin diseases, rather than cancer, and may not be adequate to prevent NHL in exposed autoworkers.

# What this study adds (100 words)

>>>>

---
# fontsize: 11pt
# mainfont: arial
output:
  # pdf_document:
  #   fig_caption: true
  #   latex_engine: lualatex
  #   includes:
  #     in_header: "~/HeadRs/Stathead.sty"
  #   keep_tex: true
  bookdown::word_document2:
    number_sections: false
    global_numbering: true
    reference_docx: "~/HeadRs/rmd_templates/style-guide-arial-11.docx"
bibliography: "`r path.expand('~/Box Sync/papers/EnvEpi.bib')`"
csl: "`r path.expand('~/HeadRs/csl/ama.csl')`"
notes-after-punctuation: true
---

@@

**Manuscript title:** Evaluating hypothetical limits on metalworking fluid exposure for reducing non-Hodgkin lymphoma incidence: An application of the hazard-extended parametric g-formula

**Authors:** Kevin T. Chen^a,b^, Sally Picciotto^b^, Ellen A. Eisen^b^

@@

**Data and computing code:** Code for reproducing analyses and reports are available on [Github](https://github.com/kvntchn/gm-nhl-ice). Data are available upon reasonable request.

**Acknowledgements:**

@@

# Abstract

(Limit: 250 words)

**Background:** Non-Hodgkin lymphoma (NHL) incidence has increased significantly since 1960 and was recently linked with workplace exposure in the United Auto Workers-General Motors (UAW-GM) Cohort Study. Here, we investigate hypothetical interventions on workplace exposure to soluble metalworking fluids (MWF) in relation to NHL risk. Interventions on entire classes of occupational exposures rather than particular agents may result in stronger lasting protections for worker health while preventing regrettable replacements.

**Methods:** We estimated counterfactual risk of NHL from 1985 to 2004 in the UAW-GM cohort of hourly autoworkers exposed to straight, soluble, and synthetic MWF at 3 Michigan plants under hypothetical interventions. We contrasted counterfactual risk under stochastic interventions on average annual exposure to soluble MWF to that under no intervention on exposure. These stochastic interventions included (1) static interventions on soluble MWF that ignore exposure to other types and (2) dynamic interventions that seek to cap total exposure to any type of MWF by intervening on exposure to soluble MWF only.

**Results:** During follow-up, 231 NHL cases occurred. Stronger hypothetical limits on average annual exposure to soluble MWF resulted in monotonically stronger reductions in NHL risk. Capping exposure to soluble MWF at 0.05 mg/m^3^ resulted in a risk ratio of 0.79 (95% CI: 0.62, 0.97).  Capping total MWF exposure at 0.05 mg/m3 by intervening only on exposure to soluble MWF resulted in a risk ratio of 0.72 (95% CI: 0.48, 1.01).

**Conclusion:** We considered realistic interventions more compatible with a population health framework than deterministic interventions and showed that stronger limits on average annual exposure to soluble MWF may confer stronger protection against NHL risk.

# What this study adds

(Limit: 100 words)

Little is known about possible links between occupational exposures and non-Hodgkin lymphoma. Existing studies are generally single time-point case-control studies with limited exposure assessment. We leveraged the large study population size and quantitative exposure assessment of the United Auto Workers-General Motors Cohort Study to estimate counterfactual risk of non-Hodgkin lymphoma under realistic interventions using a novel causal estimator capable of adjusting for time-varying confounding affected by past exposure.
